Hydra battles in Coq

### 2020-10-06 doc/hydras.tex
minor modifications

\begin{remark}
\coq's library \href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Coq.Relations.Relation\_Operators} 
contains three logically equivalent definitions of the transitive closure of a binary relation. This equivalence is proved in 
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Operators_Properties.html}{Coq.Relations.Operators\_Properties} .

@@

@@

Note the use of \coq's \texttt{sig} type in the theorem's statement, which
relates the boolean function \texttt{limitb} defined on the \texttt{T1} data-type with a constructive view of the limit of a sequence: for any $\beta<\lambda$, we can compute an item of the canonical sequence of $\lambda$ which is greater than $\beta$.

@@

@@

Let us consider a kind of accessibility problem inside $\epsilon_0$: given two ordinals $\alpha$ and $\beta$, where $\beta<\alpha<\epsilon_0$, find a \emph{path} consisting of a finite sequence $\gamma_0=\alpha,\dots,\gamma_l=\beta$,
where, for every $i<l$, $\gamma_i \not= 0$ \footnote{This condition allows us to ignore paths which end by a lot of useless $0$.} and there exists some strictly positive integer $s_i$
such that $\gamma_{i+1}=\canonseq{\gamma}{s_i}$,

@@

For instance, we have $\omega*2 \xrightarrow[2,2,2]{}\omega$, through the 
path $\omega\times 2, \omega+2,\omega+1,\omega$.

@@

For instance, we have also 
$\omega*2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega*2 \xrightarrow[{[3,14]}]{} 0$ and
$\omega*2 \xrightarrow[3,3,3,3,3,3,3,3]{} 0$.
\end{remark}

@@

Definition transition i : relation T1 :=
  match i with 0 => fun _ _ => False | S j => transition_S j end.
\end{Coqsrc}

\begin{Coqsrc}
Inductive path_to (beta: T1) : list nat -> T1 -> Prop :=
  path_to_1 : forall (i:nat) alpha , 
    i <> 0 ->
    transition i alpha beta ->
    path_to beta (i::nil) alpha
| path_to_cons : forall i alpha s gamma,
    i <> 0 ->
    transition i alpha gamma ->
    path_to beta  s gamma ->
    path_to beta  (i::s) alpha.
\end{Coqsrc}

@@

By transfinite induction on $\alpha$, we prove that for any $\beta<\alpha$, 
on can build a path from $\alpha$ to $\beta$.

@@

% ici
\noindent 
From the lemma \texttt{canonS\_LT}~\vref{lemma:canonS_LT}, we can convert any path into an inequality on ordinals (by induction on paths).

@@

In order to apply our knowledge about  ordinal numbers (less than $\epsilon_0$) to the study of hydra battles, we define an injection
from the interval $[0,\epsilon_0)$ into the type \texttt{Hydra}.

@@

The next step of our development is to extend this relationship to
the order $<$ on $[0,\epsilon_0)$ on one side, and hydra battles on the other side.

@@

We have got all the tools for proving there is no variant bounded by some $\mu<\epsilon_0$ for proving the termination   of all battles. The proof we are going to show is a proof by contradiction. It  can
 be considered as a generalization of the
proofs described in  sections~\vref{omega-case} and \vref{omega2-case}.

@@

\begin{remark}
  The hypothesis \texttt{m\_decrease} is false in general, but is satisfied by
the  \texttt{free} and \texttt{standard} kinds of battles. This trick allows to 
``factorize'' our proofs  of impossibility.
\end{remark}

@@

\begin{enumerate}
\item The following lemma is an application of \texttt{m\_ge\_generic}, since \texttt{free}
satisfies trivially the hypothesis \texttt{m\_decrease}.

@@

The tool we  use comes  once again from Ketonen and Solovay's article~\cite{KS81}. Instead of considering plain paths, i.e. sequences 
$\alpha_0=\alpha,\alpha_1,\dots,\alpha_k=\beta$ where $\alpha_{j+1}$ is equal
to $\canonseq{\alpha_j}{i_j}$ where $i_j$ is \emph{any} natural number, 
we consider various constraints on these sequences.
In particular, a path is called \emph{standard} if $i_{j+1} = i_j + 1$ for every $j<k$.
It  corresponds to a ``segment'' of some standard battles. 
Please note that the vocabulary on paths is ours, but all the concepts come really from~\cite{KS81}.

@@

We shall also use a version of \texttt{Cor12} with large inequalities.

@@

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta t1< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* rest of proof skipped *)
\end{Coqsrc}

@@

\paragraph*{}
Let us consider two ordinals  $\beta<\alpha<\epsilon_0$.  Let $p$  be some $(n+1)$-path from $\alpha$ to $\beta$.

@@

\begin{Coqsrc}
 Lemma constant_to_standard_0 : 
    {l : nat | standard_fun (S n) alpha l = beta}.

@@

\subsection{Back to hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. The proof we are going to comment can
be consulted in the module 
\url{../src/html/hydras.Hydra.Epsilon0_Needed_Std.html}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../src/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{\texttt{Lemma2\_6\_1}}   and 
\href{../src/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$.

@@

In this chapter, we try to feel how long a standard battle can be.
To be precise, for any ordinal $\alpha<\epsilon0$ and any positive integer $k$,
we give a minoration of the number of steps of a standard battle which
starts with the hydra $\iota(\alpha)$ and the replication factor $k$.

@@

\begin{Coqsrc}
Inductive L_spec : T1 -> (nat -> nat) -> Prop :=
  L_spec0 :
    forall f, (forall k, f k = k) ->  L_spec zero f
| L_spec1 : forall alpha f,
    alpha <> zero ->
    (forall k, mlarge alpha (interval (S k) (Nat.pred (f (S k))))) ->
    L_spec alpha f.
\end{Coqsrc}

@@

Let us now prove properties of any function $f$ (if any) which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plugin. Moreover, they will give us some examples of
$L_\alpha$ for small values of $\alpha$.

@@

The following lemma expresses the length of the mlarge sequences associated with finite multiples of $\omega$.

@@

To this end, we prove a generic lemma, which expresses $L_{\omega^\alpha\times i}$ in terms of $L_{\omega^\alpha}$. Note that in this correction lemma, we assume that the fonction associated with $\alpha$ is stritly monotonous and 
greater or equal than the successor function.

@@

Thus, for instance, $L_{\omega^3}(3)=L_{\omega^2\times 4}(3)$.
So, we obtain an exact expression of this number.

@@

\vspace{4pt}XS
\noindent
\emph{From Module~\href{../src/html/hydras.Epsilon0.L_alpha.html\#L_}{L\_alpha}).}

@@

Using these three lemmas as rewrite rules, we can prove more properties of the functions \texttt{L\_ $\alpha$}.

@@

\begin{Coqsrc}
Lemma H_zero_eqn : forall i, H_ Zero i = i.
Proof.   intro i; now rewrite H__equation_1.  Qed.

@@

Lemma H_lim_eqn alpha i : Limitb alpha ->
                      H_ alpha i =  H_ (Canon alpha (S i)) i.

@@

\begin{Coqsrc}
Lemma H_omega : forall k, H_ Omega k = S (2 * k)%nat.
Proof with auto with E0.
  intro k; rewrite H_lim_eqn ...
  - replace (Canon omega (S k)) with (Fin (S k)).
    + rewrite H_Fin; lia.
    +  now autorewrite with E0_rw.
Qed.
\end{Coqsrc}

@@

@@

\begin{quote}
  If $\beta<\alpha$, $F_\alpha$ dominates $F_\beta$.

($f$ is said to \emph{dominate} $g$ if $f(p)>g(p)$ for any all sufficiently large $p$).
\end{quote}

>>>>

\begin{remark}
\label{remark:transitive-closure}
\coq's library \href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Coq.Relations.Relation\_Operators} 
contains three logically equivalent definitions of the transitive closure of a binary relation. This equivalence is proved in 
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Operators_Properties.html}{Coq.Relations.Operators\_Properties} .

@@

\index{Projects}
\begin{project}
Many lemmas presented in this chapter were stated and proved before the introduction of 
the type class \texttt{ON} of ordinal notations, and in particular its  instance \texttt{Epsilon0}.
Thus definitions and lemmas refer to the type \texttt{T1} of possibly not well-formed terms.
This should be fixed in  a future version.
\end{project}

@@

@@

Note the use of \coq's \texttt{sig} type in the theorem's statement, which
relates the boolean function \texttt{limitb} defined on the \texttt{T1} data-type with a constructive view of the limit of a sequence: for any $\beta<\lambda$, we can compute an item of the canonical sequence of $\lambda$ which is greater than $\beta$.
We can also state directly that $\lambda$ is a (strict) least upper bound of the elements of its canonical sequence.

@@

Let us consider a kind of accessibility problem inside $\epsilon_0$: given two ordinals $\alpha$ and $\beta$, where $\beta<\alpha<\epsilon_0$, find a \emph{path} consisting of a finite sequence $\gamma_0=\alpha,\dots,\gamma_l=\beta$,
where, for every $i<l$, $\gamma_i \not= 0$ \footnote{This condition allows us to ignore paths which end by a lot of useless $0$s.} and there exists some strictly positive integer $s_i$
such that $\gamma_{i+1}=\canonseq{\gamma}{s_i}$,

@@

For instance, we have $\omega*2 \xrightarrow[2,2,2,4,5]{}3$, through the 
path $\langle\omega\times 2, \omega+2,\omega+1,\omega,4,3\rangle$.

@@

For instance, we have also 
$\omega*2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega*2 \xrightarrow[1,2,1,4]{} 0$ and
$\omega*2 \xrightarrow[3,3,3,3,3,3,3,3]{} 0$.
\end{remark}

@@

Definition transition i : relation T1 :=
  match i with 0 => fun _ _ => False | S j => transition_S j end.

Inductive path_to (beta: T1) : list nat -> T1 -> Prop :=
  path_to_1 : forall (i:nat) alpha , 
    i <> 0 ->
    transition i alpha beta ->
    path_to beta (i::nil) alpha
| path_to_cons : forall i alpha s gamma,
    i <> 0 ->
    transition i alpha gamma ->
    path_to beta  s gamma ->
    path_to beta  (i::s) alpha.
\end{Coqsrc}

\begin{remark}
The definition above is parameterized with the \emph{destination} of the path and indexed by the origin, hence the name \texttt{path\_to}. The rationale behind this choice is a personal preference of the developer  for the kind of eliminators generated by \coq{} in this case. The symmetric option could have been also considered (see also Remark~\vref{remark:transitive-closure}).
\end{remark}

@@

By transfinite induction on $\alpha$, we prove that for any $\beta<\alpha$, 
one can build a path from $\alpha$ to $\beta$ (in other terms, $\beta$ is accessible from $\alpha$).

@@

\noindent 
From the lemma \texttt{canonS\_LT}~\vref{lemma:canonS_LT}, we can convert any path into an inequality on ordinals (by induction on paths).

@@

In order to apply our knowledge about  ordinal numbers less than $\epsilon_0$ to the study of hydra battles, we define an injection
from the interval $[0,\epsilon_0)$ into the type \texttt{Hydra}.

@@

The next step of our development extends this relationship to
the order $<$ on $[0,\epsilon_0)$ on one side, and hydra battles on the other side.

@@

We now have  the tools for proving that  there exists no variant bounded by some $\mu<\epsilon_0$ for proving the termination   of all battles. The proof we are going to show is a proof by contradiction. It  can
 be considered as a generalization of the
proofs described in  sections~\vref{omega-case} and \vref{omega2-case}.

@@

\label{remark:m-decrease}
\begin{remark}
  The hypothesis \texttt{m\_decrease} is not provable  in general, but is satisfied by
the  \texttt{free} and \texttt{standard} kinds of battles. This trick allows to 
``factorize'' our proofs  of impossibility.
\end{remark}

@@

\begin{enumerate}
\item The following lemma is an application of \texttt{m\_ge\_generic}, since \texttt{free}
satisfies trivially the hypothesis \texttt{m\_decrease} (see page~\pageref{remark:m-decrease}).

@@

The solution of this issue comes  once again from Ketonen and Solovay's article~\cite{KS81}. Instead of considering plain paths, i.e. sequences 
$\alpha_0=\alpha,\alpha_1,\dots,\alpha_k=\beta$ where $\alpha_{j+1}$ is equal
to $\canonseq{\alpha_j}{i_j}$ where $i_j$ is \emph{any} natural number, 
we consider various constraints on these sequences.
In particular, a path is called \emph{standard} if $i_{j+1} = i_j + 1$ for every $j<k$.
It  corresponds to a ``segment'' of some standard battles. 
Please note that the vocabulary on paths is ours, but all the concepts come really from~\cite{KS81}.

@@

We  also use a version of \texttt{Cor12} with large inequalities.

@@

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta t1< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* ... *)
\end{Coqsrc}

@@

\paragraph*{}
Noq, let us consider two ordinals  $\beta<\alpha<\epsilon_0$.  Let $p$  be some $(n+1)$-path from $\alpha$ to $\beta$.

@@

\begin{Coqsrc}
 Lemma constant_to_standard_0 : 
    {l : nat | standard_fun (S n) alpha l = beta}.
 (* ... *)

@@

\subsection{Back to hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. This proof can
be consulted in the module 
\url{../src/html/hydras.Hydra.Epsilon0_Needed_Std.html}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../src/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{\texttt{Lemma2\_6\_1}}   and 
\href{../src/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$.

@@

In this chapter, we try to feel how long a standard battle can be.
To be precise, for any ordinal $\alpha<\epsilon_0$ and any positive integer $k$,
we give a minoration of the number of steps of a standard battle which
starts with the hydra $\iota(\alpha)$ and the replication factor $k$.

@@

\begin{Coqsrc}
Inductive L_spec : T1 -> (nat -> nat) -> Prop :=
| L_spec0 : forall f, (forall k, f k = k) ->  L_spec zero f
| L_spec1 : forall alpha f,
    alpha <> zero ->
    (forall k, mlarge alpha (interval (S k) (Nat.pred (f (S k))))) ->
    L_spec alpha f.
\end{Coqsrc}

@@

Let us now prove properties of any function $f$ (if any) which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plug-in. Moreover, they will give us some examples of
$L_\alpha$ for small values of $\alpha$.

@@

The following lemma expresses the length of the mlarge sequences associated with the finite multiples of $\omega$.

@@

To this end, we prove a generic lemma, which expresses $L_{\omega^\alpha\times i}$ as an iterate of $L_{\omega^\alpha}$. Note that in this lemma, we assume that the fonction associated with $\alpha$ is stritly monotonous and 
greater or equal than the successor function, and prove that $L_{\omega^\alpha\times i}$satisfies  the same properties.

@@

Thus, for instance, $L_{\omega^3}(3)=L_{\omega^2\times 4}(3)$.
Thus, we obtain an exact expression of this number.

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../src/html/hydras.Epsilon0.L_alpha.html\#L_}{L\_alpha}).}

@@

Using these three lemmas as rewrite rules, we can prove more properties of the functions \texttt{L\_$\alpha$}.

@@

\begin{Coqsrc}
Lemma H_eq1 : forall i, H_ Zero i = i.
Proof.   intro i; now rewrite H__equation_1.  Qed.

@@

Lemma H_eq3 alpha i : Limitb alpha ->
                      H_ alpha i =  H_ (Canon alpha (S i)) i.

@@

\begin{Coqsrc}
Lemma H_omega : forall k, H_ Omega k = S (2 * k)%nat.
Proof with auto with E0.
  intro k; rewrite H_eq3 ...
  - replace (Canon omega (S k)) with (Fin (S k)).
    + rewrite H_Fin; lia.
    +  now autorewrite with E0_rw.
Qed.
\end{Coqsrc}

@@

\begin{definition}
 Let $f$ zand $g$ be two arithmetic  functions; $f$ is said to \emph{dominate} $g$ if $f(p)>g(p)$ for any all sufficiently large $p$.
\end{definition}

@@

\begin{quote}
  If $\beta<\alpha$, $F_\alpha$ dominates $F_\beta$.
\end{quote}
